### 2018-09-23 functionalities_table.tex
work++

\begin{document}
\pagenumbering{gobble}
\begin{tabular}{|l|l|l|}
	\hline
	Package & Role & Main functions \\ \hline \hline
	\multicolumn{ 1}{|c|}{behavr} & utilities & \texttt{behavr()}, \texttt{meta()},\texttt{setbehavr()},\texttt{setmeta()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & metadata operations \textbf{see S1 Fig.} & \texttt{xmd()}, \texttt{rejoin()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & merge and check several tables & \texttt{bind\_behavr\_list()}, \texttt{stitch\_on()} \\ \hline
	\multicolumn{ 1}{|c|}{ggetho} & interface to ggplot (\textbf{see main text}) & \texttt{ggetho()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|c|}{} & tile and ``bar tile'' plots (\textbf{\emph{e.g.} Fig. 3 and 4A}) & \texttt{stat\_tile\_etho}, \texttt{stat\_bar\_tile\_etho()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & population trends as lines and error bars & \texttt{stat\_pop\_etho()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & circadian biology utilities (\textbf{\emph{e.g.} Fig. 4 and 5A}) & \texttt{stat\_ld\_annotations()}, \texttt{geom\_peak()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & smart time scales (auto-adjust labels and ticks) & \texttt{scale\_x\_days()}, \texttt{scale\_x\_hours()}, \dots\\ \hline
	\multicolumn{ 1}{|c|}{damr} & ``link'' metadata to DAM files (\textbf{see main text}) & \texttt{link\_dam\_metadata()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & \makecell[vl]{``load'' relevant DAM (2 or 5) data as behavr.\\Detects daylight-saving bugs and such} & \texttt{load\_dam()} \\ \hline
	\multicolumn{ 1}{|c|}{scopr} & ``link'' metadata to ethoscope files & \texttt{link\_ethoscope\_metadata()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & ``load'' relevant ethoscope data as behavr & \texttt{load\_ethoscope()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & utilities (list files, fetch remote data, ) & \texttt{experiment\_info()}, \texttt{list\_result\_files()}, \dots \\ \hline
	\multicolumn{ 1}{|c|}{zeitgebr} & compute periodograms (AC, $\chi^2$,LS,Fourrier) & \small{\texttt{*\_periodogram()} (\texttt{*} $\in$ \texttt{\{ac,chi\_sqr,ls,fourrier\}})}\\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & preprocess and standardise methods & \texttt{periodogram()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & find peak periods in periodograms (\textbf{\emph{e.g.} Fig. 4 and 5A}) & \texttt{find\_peaks()} \\ \hline
	\multicolumn{ 1}{|c|}{sleepr} & motion thresholding/velocity detection & \texttt{motion\_detectors()}  \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & sleep definition using N minute rule & \texttt{sleep\_annotation()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & data curation: dead animals scored as asleep & \texttt{curate\_dead\_animals()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & discrete behavioural state bout architecture & \texttt{bout\_analysis()} \\ \hline
\end{tabular}

>>>>

\begin{document}
\pagenumbering{gobble}
\begin{tabular}{|l|l|l|}
	\hline
	Package & Role & Main functions \\ \hline \hline
	\multicolumn{ 1}{|c|}{behavr} & utilities & \texttt{behavr()}, \texttt{meta()},\texttt{setbehavr()},\texttt{setmeta()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & metadata operations \textbf{see S1 Fig.} & \texttt{xmd()}, \texttt{rejoin()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & merge and check several tables & \texttt{bind\_behavr\_list()}, \texttt{stitch\_on()} \\ \hline
	\multicolumn{ 1}{|c|}{ggetho} & interface to ggplot (\textbf{see main text}) & \texttt{ggetho()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|c|}{} & tile and ``bar tile'' plots (\textbf{\emph{e.g.} Fig. 3 and 4A}) & \texttt{stat\_tile\_etho}, \texttt{stat\_bar\_tile\_etho()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & population trends as lines and error bars & \texttt{stat\_pop\_etho()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & circadian biology utilities (\textbf{\emph{e.g.} Fig. 4 and 5A}) & \texttt{stat\_ld\_annotations()}, \texttt{geom\_peak()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & visualise spectrograms (\textbf{\emph{e.g.} CWT, Fig. 6}) & \texttt{ggspectro()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & smart time scales (auto-adjust labels and ticks) & \texttt{scale\_x\_days()}, \texttt{scale\_x\_hours()}, \dots\\ \hline
	\multicolumn{ 1}{|c|}{damr} & ``link'' metadata to DAM files (\textbf{see main text}) & \texttt{link\_dam\_metadata()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & \makecell[vl]{``load'' relevant DAM (2 or 5) data as behavr.\\Detects daylight-saving bugs and such} & \texttt{load\_dam()} \\ \hline
	\multicolumn{ 1}{|c|}{scopr} & ``link'' metadata to ethoscope files & \texttt{link\_ethoscope\_metadata()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & ``load'' relevant ethoscope data as behavr & \texttt{load\_ethoscope()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & utilities (list files, fetch remote data, ) & \texttt{experiment\_info()}, \texttt{list\_result\_files()}, \dots \\ \hline
	\multicolumn{ 1}{|c|}{zeitgebr} & compute periodograms (CWT, AC, $\chi^2$,LS,Fourrier) & \small{\texttt{*\_periodogram()} (\texttt{*} $\in$ \texttt{\{cwt, ac,chi\_sqr,ls,fourrier\}})}\\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & preprocess and standardise methods & \texttt{periodogram()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & find peak periods in periodograms (\textbf{\emph{e.g.} Fig. 4 and 5A}) & \texttt{find\_peaks()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & compute CWT spectrograms & \texttt{spectrogram()} \\ \hline	
	\multicolumn{ 1}{|c|}{sleepr} & motion thresholding/velocity detection & \texttt{motion\_detectors()}  \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & sleep definition using N-minute rule & \texttt{sleep\_annotation()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & data curation: dead animals scored as asleep & \texttt{curate\_dead\_animals()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & discrete behavioural state architecture & \texttt{bout\_analysis()} \\ \hline
\end{tabular}

### 2018-09-23 manuscript.Rnw
work++

\section*{Design and Implementation}
\texttt{rethomics} is implemented in \texttt{R}\cite{r_core_team_r_2017}
as a collection of packages (Fig~\ref{fig:fig-1}).
A summary of the functionality of all packages is represented in a supplementary table (\nameref{S1-Table}).
Such modular architecture follows the model of modern frameworks such as the \texttt{tidyverse}\cite{wickham_tidyverse_2017}, which results in increased testability, maintainability and adaptability.
In this model, each task of the analysis workflow (\emph{i.e.} data import, manipulation and visualisation) is handled by a different package, and new ones can be designed to suit specific needs.
At the core of \texttt{rethomics} lies the \texttt{behavr} table, a structure used to store large amounts of data (\emph{e.g.} position and activity) and metadata (\emph{e.g.} treatment and genotype) in a unique \texttt{data.table}-derived object\cite{dowle_data.table_2017}.
Any input package will import experimental data as a \texttt{behavr} table which can, in turn, be analysed and visualised regardless of the original input platform.
Numerical results and plots are standard objects that can, therefore, be further analysed inside the wide \texttt{R} package ecosystem.

@@

\section*{Discussion}
The framework  we presented in this article describes a general computational set of tools to store, manipulate and analyse behavioural data.
We showed that \texttt{rethomics} can readily be used to analyse specifically circadian data. As such, it already fills a clear gap in the sleep and circadian community by providing an open, extensible and programmatic toolbox as opposed to the rigid array of GUI tools which cannot interoperate.
In particular, within the \texttt{R} environment, \texttt{rethomic} features a streamlined workflow (Fig.~\ref{fig:fig-1}):
data loading; quality control (Fig.~\ref{fig:fig-3}); per individual visualisation (Fig.~\ref{fig:fig-4}A) and computation (Fig.~\ref{fig:fig-4}B); group averages and statistics (Fig.~\ref{fig:fig-5}), ultimately resulting in publication-quality figures.
In addition, to the canonical analysis we demonstrated, the richness of the \texttt{R} environment renders rethomics readily available for exploratory work such as the use of continuous wavelet transform to study the modulation of walking pace(Fig.~\ref{fig:fig-6}).

Altogether, our tool provides a needed opportunity for the community to share analytic and visualisation tools whilst embracing the diversity of recording equipment. By design, rethomics is very flexible and not limited to the study of circadian rhythms.
In fact, \texttt{behavr} tables (Fig.~\ref{fig:fig-2}) are a general solution to link data to metadata that come with no assumption regarding the type and number of variables stored. Therefore, we anticipate it will apply to  a variety contexts and scales.

>>>>

\section*{Design and Implementation}
\texttt{rethomics} is implemented in \texttt{R}\cite{r_core_team_r_2017}
as a collection of packages (Fig~\ref{fig:fig-1}).
A summary of the functionalities of all packages is represented in a supplementary table (\nameref{S1-Table}).
Such modular architecture follows the model of modern frameworks such as the \texttt{tidyverse}\cite{wickham_tidyverse_2017}, which results in increased testability, maintainability and adaptability.
In this model, each task of the analysis workflow (\emph{i.e.} data import, manipulation and visualisation) is handled by a different package, and new ones can be designed to suit specific needs.
At the core of \texttt{rethomics} lies the \texttt{behavr} table, a structure used to store large amounts of data (\emph{e.g.} position and activity) and metadata (\emph{e.g.} treatment and genotype) in a unique \texttt{data.table}-derived object\cite{dowle_data.table_2017}.
Any input package will import experimental data as a \texttt{behavr} table which can, in turn, be analysed and visualised regardless of the original input platform.
Numerical results and plots are standard objects that can, therefore, be further analysed inside the wide \texttt{R} package ecosystem.

@@

\section*{Discussion}
The framework  we presented in this article describes a general computational set of tools to store, manipulate and analyse behavioural data.
We showed that \texttt{rethomics} can readily be used to analyse circadian data. As such, it already fills a clear gap in the sleep and circadian community by providing an open, extensible and programmatic toolbox as opposed to a rigid array of GUI tools which cannot interoperate.
In particular, within the \texttt{R} environment, \texttt{rethomic} features a streamlined workflow (Fig.~\ref{fig:fig-1}):
data loading; quality control (Fig.~\ref{fig:fig-3}); per individual visualisation (Fig.~\ref{fig:fig-4}A) and computation (Fig.~\ref{fig:fig-4}B); group averages and statistical analysis (Fig.~\ref{fig:fig-5}), ultimately resulting in publication-quality figures.
In addition to the canonical analysis we presented in this study, the richness of the \texttt{R} environment renders rethomics particularly interesting for exploratory work such as the use of continuous wavelet transform to study the modulation of walking pace(Fig.~\ref{fig:fig-6}).

Altogether, our tool provides a needed opportunity for the community to share analytic and visualisation solutions whilst embracing the diversity of recording equipment. By design, rethomics is very flexible and not limited to the study of circadian rhythms.
In fact, \texttt{behavr} tables (Fig.~\ref{fig:fig-2}) are a general structure to link data to metadata that come with no assumption regarding the type and number of variables stored.
Therefore, we anticipate it will apply to  a variety of contexts and scales.
For instance, in the last decade, multiple devices have been developed -- either commercially or academically -- to record longitudinal data on several animal models (\emph{i.e.} feeding, metabolic activity, heart rate, position and so on). In collaboration with the teams providing of these methods, we envisage the development of device-specific packages that will load data into \texttt{rethomics}, making visualisation and analysis immediately available. 
For instance, the developer any behaviour recording method can simply write a small package to read their data in a \texttt{behavr} table and benefit from the functionalities available.
Furthermore, such an approach becomes particularily useful in order to compare methods or study behaviours that are recorded by different devices, on the same individuals. Indeed, in these contexts, it becomes apparent that analysis must be unified at some level.

In conclusion, \texttt{rethomics} already delivers state-of-the-art tools to study circadian rhythm using conventional tools, but was designed in an inclusive and extensible manner. As such, if fills a unique niche in a community that is becoming increasingly quantitative.

### 2018-09-23 manuscript.tex
work++

To illustrate how \texttt{rethomic} integrates with other packages and render such non-mainstream analysis possible,
we performed a wavelet analysis of the position of 80 naive fruit flies (40 females and 40 males) in their glass tubes.
We used the package \texttt{scopr}, part of \texttt{rethomics}, to load five days of ethoscope positional data, which we sampled at 0.1~Hz.
Our variable of interest is the position of animals in their tube (from the food end, $Position = 0$,  to the cotton end, $Position = 1$).
Fig~\ref{fig:fig-6}A-C shows the raw position data at two different scales for two representative animals.

@@

In order to compute CWT, we used the \texttt{WaveletComp} package\cite{schmidbauer_waveletcomp_2018}.
We then averaged the result of the five consecutive days in the time/period domain over one circadian day both for the two representative animals (Fig~\ref{fig:fig-6}D)
and for the population(Fig~\ref{fig:fig-6}E).

>>>>

To illustrate how \texttt{rethomic} integrates with other packages and render such non-mainstream analysis possible, 
we developped, as part of the \texttt{zeitgebr} package, utilities to perform wavelet analysis, which we then applied to the position of 80 naive fruit flies (40 females and 40 males) in their glass tubes.
We used the package \texttt{scopr}, to load five days of ethoscope positional data, which we sampled at 0.1~Hz.
Our variable of interest is the position of animals in their tube (from the food end, $Position = 0$,  to the cotton end, $Position = 1$).
Fig~\ref{fig:fig-6}A-C shows the raw position data at two different scales for two representative animals.

@@

In order to compute CWT, we interfaced the \texttt{WaveletComp} package\cite{schmidbauer_waveletcomp_2018} in \texttt{rethomics}. 
This interface allowed us to average the result of the five consecutive days in the time/period domain over one circadian day both for the two representative animals (Fig~\ref{fig:fig-6}D)
and for the population(Fig~\ref{fig:fig-6}E).
